Fully Automatic Loop-free Program Synthesis by Bounded Model Checking

### 2013-10-09 paper/paper.tex
Added technical contributions. Minor changes

Our work represents a substantial refinement of {\sc brahma}~\cite{brahma},
which is the closest work to ours.  {\sc Brahma} also focuses on
synthesizing loop-free machine-integer programs.  In~\cite{brahma}, the
authors characterise the program synthesis problem as a second-order
$\exists \forall$ formula and uses the CEGIS algorithm,
introduced in~\cite{lezama-thesis}, for solving
these formulae.  This algorithm reduces the satisfiability of the $\exists
\forall$ formula to the satisfiability of several first-order formulae, each
of which can be decided using an off-the-shelf SMT solver.  {\sc Brahma}
takes as input a specification (given as an SMT formula) and a library of
components (one component is more or less a single machine instruction).  It
then finds a permutation of the component library that satisfies the
specification.  This permutation is the final synthesised program.  In the
event that the library does not contain enough components to synthesise a
correct program, the user is told that the library is too small and is
must add more components until the synthesis succeeds.  Each round of
synthesis amounts to solving an $\exists \forall$ formula that is quadratic
in the size of the component library.

@@

Program synthesis is strongly related to
superoptimisation~\cite{superoptimisation}, which is the problem of taking a
segment of code, usually written in C, and finding a minimal program which
has the same functionality.  The {\sc aha}~\cite{aha} tool and the GNU
superoptimiser~\cite{gnu-superoptimiser} both perform superoptimisation of C
programs, both by enumerating possible programs and testing them on suitable
inputs.  This search can be compiled down to native code using an optimising
compiler, so for short programs the method is extremely fast.  However, this
method is impractical for synthesising longer programs because of the
exponential blowup of the search space.  We incorporate the core idea of
this technique by using explicit-state model checking in parallel with
symbolic model checking.

@@

@@

When the first-order fragment of the logic is efficiently decidable and $P$ can be
expressed as a ground term of the logic, we can use a refinement loop to check the
validity of the synthesis formula as shown in Fig.~\ref{fig:abstract-refinement} and
Fig.~\ref{fig:abstract-refinement-code}.  The algorithm is divided into two
components: {\sc synth} and {\sc verif}.  We track a finite (small) set of inputs
and {\sc synth} synthesises a candidate program that is correct on just those inputs.
{\sc Verif} then tries to verify the candidate program by searching for an input on
which the candidate program does not satisfy the specification.  If no such input
can be found, the candidate program is correct.  Otherwise, the new input is added
to the set of inputs we must check in {\sc synth}.  If the input set is finite, this
procedure is guaranteed to terminate.

>>>>

Our work represents a substantial refinement of {\sc brahma}~\cite{brahma},
which is the closest work to ours.  {\sc Brahma} also focuses on
synthesizing loop-free machine-integer programs.  In~\cite{brahma}, the
authors characterise the program synthesis problem as a second-order
$\exists \forall$ formula and uses a variation of the CEGIS algorithm,
introduced in~\cite{lezama-thesis}, for solving
these formulae.  This algorithm reduces the satisfiability of the $\exists
\forall$ formula to the satisfiability of several first-order formulae, each
of which can be decided using an off-the-shelf SMT solver.  {\sc Brahma}
takes as input a specification (given as an SMT formula) and a library of
components (one component is more or less a single machine instruction).  It
then finds a permutation of the component library that satisfies the
specification.  This permutation is the final synthesised program.  In the
event that the library does not contain enough components to synthesise a
correct program, the user is told that the library is too small and is
must add more components until the synthesis succeeds.  Each round of
synthesis amounts to solving an $\exists \forall$ formula that is quadratic
in the size of the component library.

@@

Program synthesis is closely related to
superoptimisation~\cite{superoptimisation}, which is the problem of taking a
segment of code, usually written in C, and finding a minimal program which
has the same functionality.  The {\sc aha}~\cite{aha} tool and the GNU
superoptimiser~\cite{gnu-superoptimiser} both perform superoptimisation of C
programs, both by enumerating possible programs and testing them on suitable
inputs.  This search can be compiled down to native code using an optimising
compiler, so for short programs the method is extremely fast.  However, this
method is impractical for synthesising longer programs because of the
exponential blowup of the search space.  We incorporate the core idea of
this technique by using explicit-state model checking in parallel with
symbolic model checking.

@@

\section{Technical Contributions}
Our approach generates an $\exists \forall$ formula which is linear
in the size of the synthesised program.  This improves over the
enoding of~\cite{brahma}, which generates a $\exists \forall$ formula
that is quadratic in the size of the synthesised program.

We introduce a novel parameterisation of the programming language
used to express our synthesised programs.
This parameterisation allows us to efficiently explore the program space
without relying on human guidance.

Our tool is the first we are aware of that is able to effectively
synthesise floating point programs.  We demonstrate this by
synthesising {\sc Fast2Sum} using Knuth's {\sc 2Sum}~\cite{taocp2} as
a specification.

@@

When the first-order fragment of the logic is efficiently decidable and $P$ can be
expressed as a ground term of the logic, we can use
Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis, sketch} to
check the validity of the synthesis formula.  The core of the CEGIS algorithm is
a refinement loop to check the
validity of the synthesis formula as shown in Fig.~\ref{fig:abstract-refinement} and
Fig.~\ref{fig:abstract-refinement-code}.  The algorithm is divided into two
components: {\sc synth} and {\sc verif}.  We track a finite (small) set of inputs
and {\sc synth} synthesises a candidate program that is correct on just those inputs.
{\sc Verif} then tries to verify the candidate program by searching for an input on
which the candidate program does not satisfy the specification.  If no such input
can be found, the candidate program is correct.  Otherwise, the new input is added
to the set of inputs we must check in {\sc synth}.  If the input set is finite, this
procedure is guaranteed to terminate.
